---
title: "Transform Data"
output: html_document
---

```{r setup}
library(tidyverse)
library(gapminder)
```

## gapminder

```{r}
gapminder
```


## Your Turn 1

Alter the code to select just the `pop` column:

```{r}
select(gapminder, year, lifeExp)
```

## Your Turn 1: Bonus

Finish early? Here's a bonus question: Which of these is NOT a way to select the `country` and `continent` columns together?

```{r}
select(gapminder, -c(year, lifeExp, pop, gdpPercap))
select(gapminder, country:continent)
select(gapminder, starts_with("c"))
select(gapminder, ends_with("t"))
```

## Your Turn 2

Show:

* All of the rows where `pop` is greater than or equal to 100000  
* All of the rows for El Salvador  
* All of the rows that have a missing value for `year`  (no need to edit this code)

```{r}


filter(gapminder, is.na(year))
```

## Your Turn 3

Use Boolean operators to alter the code below to return only the rows that contain:

* El Salvador  
* Countries that had populations over 100000 in 1960 or earlier

```{r}
filter(gapminder, country == "El Salvador" | country == "Oman")
filter(______, _______)
```

## Your Turn 4

Use `%>%` to write a sequence of functions that: 

1. Filter only countries that are in the continent of Oceania.
2. Select the `country`, `year` and `lifeExp` columns  
3. Arrange the results so that the highest life expectancy is at the top.

```{r}

```


## Your Turn 5

Use `summarise()` to compute three statistics about the gapminder data set:

1. The first (`min()`) year in the data
2. The last (`max()`) year in the data  
3. The total number of observations (`n()`) and the total number of unique countries in the data (`n_distinct()`)

```{r}

```

## Your Turn 6

Use grouping to calculate the mean life expectancy for each continent and year. Call the mean life expectancy variable `mean_le`. Plot the life expectancy over time (no need to change the plot code).

```{r}
gapminder %>% 
  __________ %>% 
  __________ %>% 
  ggplot(aes(x = year, y = mean_le, col = continent)) +
    geom_line() + 
    scale_color_manual(values = continent_colors)
```

## Your Turn 6: Bonus

Finish early? Try this bonus exercise:

Extract the rows where `continent == "Europe"`. Then use `group_by()` to group by country. Finally, use `summarize()` to compute:

1. The total number of observations for each country in Europe
2. The lowest observed life expectancy for each country 

```{r}

```

***

# Take aways

* Extract variables with `select()`  
* Extract rows with `filter()`  
* Arrange rows, with `arrange()`  

* Make tables of summaries with `summarise()`  
* Make new variables, with `mutate()`  
* Do groupwise operations with `group_by()`

* Connect operations with `%>%`  

What else can you do with dplyr?

* Use `left_join()`, `right_join()`, `full_join()`, or `inner_join()` to join datasets
* Use `semi_join()` or `anti_join()` to filter datasets against each other

* `dplyr` works seamlessly with databases


