---
title: "Visualizing epidemiologic data in R and RStudio"
author: "Corinne Riddell"
date: "November 19, 2020"
output: slidy_presentation
---

### Learning objectives:

1. To put to use the `dplyr` commands from the first session
2. To make beautiful plots using the `ggplot2` package

### Life expectancy in the United States by race and gender, 1969-2013 

These data are partial results from a study that I did on the difference in life 
expectancy between non-Hispanic Black and White men and women in the United 
States over time. 

A subset of the results have been stored in the data/ folder as a CSV file.

Do you remember which function to use to import CSV data into R?

### `readr`'s `read_csv()` to import these data

```{r import-csv-data}
library(readr) #readr is part of the tidyverse
le_data <- read_csv("./data/Life-expectancy-by-state-long.csv")
```

### Five functions to get to know your dataset

Function 1
```{r first-lines}
head(le_data)
```

### Five functions to get to know your dataset

Function 2
```{r dimensions}
dim(le_data)
```

### Five functions to get to know your dataset

Function 3
```{r names}
names(le_data)
```

### Five functions to get to know your dataset

Function 4
```{r structure}
str(le_data)
```

### Five functions to get to know your dataset

Function 5
```{r view, eval = F}
View(le_data)
```

To RStudio!


### Summary: Five functions to get to know your dataset

1. `head()`: prints the first 6 lines of a data frame
2. `dim()`: prints the # rows and # columns
3. `names()`: prints the variable names
4. `str()`: shows the type of each variable and some values
5. `View()`: opens the viewer pane in RStudio

### Life expectancy for White men in California

Make a scatter plot of the life expectancy for White men in California
over time.

Since the dataset contains 39 states across two genders and two races, first use
a function to subset the data to contain only White men in California.

Which function from Malcolm's lesson do we need?

### `dplyr`'s `filter()` to select a subset of rows

```{r wh-cali-dataset, message=F}
library(dplyr)
wm_cali <- le_data %>% filter(state == "California", 
                              sex == "Male", 
                              race == "white")

#this is equivalent:
wm_cali <- le_data %>% filter(state == "California" & sex == "Male" & race == "white")
```

### First step to building a `ggplot()`: set up a canvas

- The line of code specified the `data` set and what goes on the `x` and `y` axes

```{r ggplot-canvas, fig.width=5.5, fig.height=4}
library(ggplot2)
ggplot(data = wm_cali, aes(x = year, y = LE))
```

### Second step to building a `ggplot()`: tell `ggplot` how to plot the data

```{r ggplot-geom_point, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_point()
```

- `geom_point()` tells ggplot to use points to plot these data

### `labs()` to add a title, a caption, and modify x and y axes titles

```{r ggplot-change-titles, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_point() +
  labs(title = "Life expectancy in White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### `col` controls the color of geom_point()

```{r ggplot-change-col, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_point(col = "blue") +
  labs(title = "Life expectancy in White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### `size` controls the size of geom_point()

```{r ggplot-change-col-size, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_point(col = "blue", size = 4) +
  labs(title = "Life expectancy in White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### Line plot rather than scatter plot

What if we wanted to make these data into a line plot instead. What part of the 
code should change?

```{r ggplot-previous-code, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + 
  geom_point(col = "blue", size = 4) +
  labs(title = "Life expectancy in White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### `geom_line()` to make a line plot

```{r ggplot-geom_line, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_line(col = "blue") +
  labs(title = "Life expectancy in White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### Life expectancy for White and Black men in California

What do we need to change to make a separate line for both Black and White men?

### First, update the `filter()`

```{r create-new-dataset}
wbm_cali <- le_data %>% filter(state == "California",
                               sex == "Male")
```

### Look at the previous code and output first:

```{r ggplot-geom_line-dup, fig.width=5.5, fig.height=4}
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_line(col = "blue") +
  labs(title = "Life expectancy in White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### And change it to *link* color to race

```{r ggplot-geom_line-aes, fig.width=5.5, fig.height=4}
ggplot(data = wbm_cali, aes(x = year, y = LE)) + geom_line(aes(col = race)) +
  labs(title = "Life expectancy in Black and White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### Always use the aes() function to link a plot feature to a variable in your data frame

The operative word is *link*. Whenever you want to link something about how 
the plot looks to a variable in the data frame, you need to *link* these items
inside the `aes()` function:

```{r}
ggplot(data = wbm_cali, aes(x = year, y = LE)) + geom_line(aes(col = race)) +
  labs(title = "Life expectancy in Black and White men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### The `aes()` function

* What else was added to the plot when you used the `aes()` function?

### The `aes()` function

* What else was added to the plot when you used the `aes()` function?
    * A legend was added showing the link between the line color and the data 
    frame's race variable

### What if we also wanted to look at women?

### What if we also wanted to look at women?
```{r cali-dataset, fig.width=5.5, fig.height=4}
cali_data <- le_data %>% filter(state == "California")
```

### What is wrong with this plot?

```{r ggplot-whoopsies, fig.width=5.5, fig.height=4}
ggplot(data = cali_data, aes(x = year, y = LE)) + geom_line(aes(col = race)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### Use `lty()` to **link** line type to sex

```{r ggplot-california, fig.width=5.5, fig.height=4}
ggplot(data = cali_data, aes(x = year, y = LE)) + geom_line(aes(col = race, lty = sex)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

### Use `facet_wrap()` to make separate plots for a specified variable

```{r ggplot-california-facet, fig.width=7.5, fig.height=4}
ggplot(data = cali_data, aes(x = year, y = LE)) + 
  geom_line(aes(col = race, lty = sex)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)") +
  facet_wrap(~ sex) 
```

### Compare two states

How do we update the `filter` to include data from California and New York?

### Compare two states

```{r two-states}
updated_data <- le_data %>% filter(state %in% c("California", "New York"))
```

### Let's write the code together

```{r}
#to fill in during class
```

### Let's write the code together

```{r ggplot-facet_wrap, fig.width=7, fig.height=7}
ggplot(data = updated_data, aes(x = year, y = LE)) + 
  geom_line(aes(col = race, lty = sex)) +
  labs(title = "Life expectancy in California and New York, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)") +
  facet_grid(state ~ sex)
```

### Question

What is the difference between `facet_wrap()` and `facet_grid()`?

### So far

- `geom_point()` to make scatter plots
- `geom_line()` to make line plots
- `col = "blue"`, `size = 2`, `lty = 2`, to change color, size and line type of the `geom`
- `aes(col = race)` to *link* color to race
- `aes(lty = sex)` to *link* line type to sex
- `facet_wrap(~ var1)` to make separate plots for different levels of one variable
- `facet_grid(var1 ~ var2)` to make separate plots for combinations of levels of two variables

### What if we wanted to make a histogram...

...of life expectancy of White men in 2013?

Before you code, try and visualize what the histogram will show

- What is on the x axis? 
- What is on the y axis? 

### Update the `filter`

```{r filter-2013-men-white}
wm_data <- le_data %>% filter(year == 2013, sex == "Male", race == "white")
```

### `geom_histogram()` to make histograms

```{r ggplot_geom_histogram, fig.width=5.5, fig.height=4}
ggplot(dat = wm_data, aes(x = LE)) + geom_histogram()
```

### use `fill` to change the *fill* of the histogram and `binwidth` to specify the bin width 


```{r ggplot_geom_histogram-bin-width, fig.width=5.5, fig.height=4}
wm_data <- le_data %>% filter(year == 2013, sex == "Male", race == "white")

ggplot(dat = wm_data, aes(x = LE)) + 
  geom_histogram(binwidth = 1, col = "white", fill = "forest green") 
```

### Apply some of our new skills

```{r ggplot_geom_histogram-more, fig.width=5.5, fig.height=4}
data_2013 <- le_data %>% filter(year == 2013)

ggplot(dat = data_2013, aes(x = LE)) + 
  geom_histogram(binwidth = 1, col = "white", aes(fill = sex)) + 
  facet_grid(race ~ sex)
```

### Recap: What functions did we learn?

1. `ggplot()`
    - `geom_scatter()`
    - `geom_line()`
    - `geom_histogram()`
    - `aes()` to link aesthetics to variables in our data frame
    - `facet_wrap(~ var1)`, `facet_grid(var1 ~ var2)`
    - `labs(title = "Main", y = "y axis", x = "x axis", caption = "below plot")`

### Recap: What arguments were useful?

2. `ggplot()`
    - `col`
    - `size`
    - `lty`
    
### We only skimmed the surface!

- You now have a sense of how `ggplot` works, but you might be itching to learn more.
    - [Kieran Healy's data visualization book](https://www.amazon.com/gp/product/0691181624/ref=as_li_tl?ie=UTF8&tag=kieranhealysw-20&camp=1789&creative=9325&linkCode=as2&creativeASIN=0691181624&linkId=16d53b3cc1ec3bc3aac60b27c29b92e8)
    - [RStudio ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

### Where to ask ggplot2 questions

- [The RStudio community page](https://community.rstudio.com/)
- [Stack Overflow](https://stackoverflow.com/)
- On Twitter using the #rstats hashtag
