---
title: "Reproduce Reports"
output: html_document
params: 
  continent: Americas
bibliography: "misc/citations.bib"
---

Chunks named "setup" are a little special. They will always run first in R Notebooks and interactively if they haven't been run yet.

```{r setup, echo = FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(gapminder)
```

## Your turn 1

Create a new R Markdown file. Go to File > New File > R Markdown. Press OK. Save the file and press the "Knit" button above.

## Your turn 2

Create a code chunk. You can type it in manually, use the keyboard short-cut (Cmd/Ctrl + Option/Alt + I), or use the "Insert" button above. Put the following code in it:

gapminder %>% 
  slice(1:5) %>% 
  knitr::kable()
  
Knit the document

## Your turn 2: Bonus
  
Add `echo = FALSE` to the code chunk above and re-knit. What's different?

## Your turn 3

1. Remove `eval = FALSE` so that R Markdown evaluates the code.
2. Use `summarize()` and `n_distinct()` to get the the number of unique years in gapminder and save the results as `n_years`.
3. Use inline code to describe the data set in the text below the code chunk and re-knit.

```{r, eval = FALSE}
___ ___ gapminder %>% 
  _________ %>% 
  # pull() extracts a column as a vector
  # if no column is called, it uses the most recently created
  pull()
```

Gapminder observed data over _________ different years.


## Your turn 4

1. Set figure chunk options such as `dpi`, `fig.width`, and `fig.height`. Run `knitr::opts_chunk$get()` in the console to see the defaults.
2. Change the YAML header above from `output: html_document` to another output type like `pdf_document` or `word_document`.
3. Add your name to the YAML header using `author: Your Name`.

```{r}
gapminder %>% 
  filter(continent != "Oceania") %>% 
  ggplot(aes(x = year, y = lifeExp, group = country, color = country)) +
  geom_line(lwd = 1, show.legend = FALSE) + 
  facet_wrap(~ continent) +
  # `country_colors` is an object included in gapminder
  scale_color_manual(values = country_colors) + 
  theme_minimal(14) + 
  theme(strip.text = element_text(size = rel(1.1)))
```

# Take aways

* R Markdown has simple syntax that's easy to use
* Your analysis and narrative are easy to intertwine in R Markdown
* You can re-run your analysis with different inputs and outputs
* Code chunks provide a variety of options to change the resulting report
* Use YAML to set options for R Markdown
* R Markdown supports a powerful citation system. Find more CSL styles at https://github.com/citation-style-language/styles

# Bonus Material

Want a little more R Markdown? Try this bonus material

## Learn Markdown Syntax
To better learn markdown syntax, try the ten-twenty minute tutorial at https://commonmark.org/help/tutorial. 

## Create a parameterized report

Read the chapter on parameterized R Markdown reports in the R Markdown book: https://bookdown.org/yihui/rmarkdown/parameterized-reports.html

1. Change the `params` option in the YAML header to use a different continent. Re-knit

```{r}
gapminder %>% 
  filter(continent == params$continent) %>% 
  ggplot(aes(x = year, y = lifeExp, group = country, color = country)) +
  geom_line(lwd = 1, show.legend = FALSE) + 
  scale_color_manual(values = country_colors) +
  theme_minimal(14) + 
  theme(strip.text = element_text(size = rel(1.1))) +
  ggtitle(paste("Continent:", params$continent))
```

## Create citations

Read the chapter on citations and bibliographies from R Markdown for Scientists: https://rmd4sci.njtierney.com/citing-articles-bibliography-styles.html

1. Cite the Causal Inference book in text below in the format ```[@citation-label]``` (without the tick marks; you only need those to tell R Markdown not to evaluate the this). The label for the citation is `hernan_causal_2019`
2. Add the American Journal of Epidemiology CSL to the YAML using `csl: misc/aje.csl`

We calculated the average treatment effect using g-computation ________.


